---
title: "Cache Dependency Graph — Deep Dive"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cache Dependency Graph — Deep Dive}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 5
)
```

cacheR automatically tracks **which cached function called which** and **which files each depends on**. This vignette explores the full graph API: building, inspecting, persisting, and visualising the dependency DAG.

## 1. Building a Pipeline

When a cached function calls another cached function, cacheR records the parent → child relationship automatically.

```{r}
library(cacheR)

cache_dir <- file.path(tempdir(), "graph_demo")
if (dir.exists(cache_dir)) unlink(cache_dir, recursive = TRUE)
cacheTree_reset()

load_counts <- cacheFile(cache_dir, backend = "rds") %@% function(path) {
  list(genes = c("TP53", "BRCA1", "EGFR"), counts = c(100, 250, 80))
}

normalize <- cacheFile(cache_dir, backend = "rds") %@% function(counts) {
  total <- sum(counts$counts)
  setNames(counts$counts / total, counts$genes)
}

top_genes <- cacheFile(cache_dir, backend = "rds") %@% function(normed, n = 2) {
  head(sort(normed, decreasing = TRUE), n)
}

pipeline <- cacheFile(cache_dir, backend = "rds") %@% function(input_path, n = 2) {
  raw <- load_counts(input_path)
  normed <- normalize(raw)
  top_genes(normed, n = n)
}

result <- pipeline("samples/counts.csv")
print(result)
```

## 2. Inspecting Nodes

Each node stores: function name, hash, parent/child links, tracked files, and cache file path.

```{r}
nodes <- cacheTree_nodes()
cat(sprintf("Total nodes: %d\n\n", length(nodes)))

for (nid in names(nodes)) {
  nd <- nodes[[nid]]
  cat(sprintf("Node: %s\n", nd$fname))
  cat(sprintf("  ID:       %s\n", nid))
  cat(sprintf("  Parents:  %s\n", paste(nd$parents, collapse = ", ")))
  cat(sprintf("  Children: %s\n", paste(nd$children, collapse = ", ")))
  cat(sprintf("  Files:    %s\n", paste(nd$files, collapse = ", ")))
  cat("\n")
}
```

## 3. File Tracking with `track_file()`

`track_file(path)` registers an explicit file dependency on the current graph node.

```{r}
cacheTree_reset()
cache_dir2 <- file.path(tempdir(), "graph_demo_files")
if (dir.exists(cache_dir2)) unlink(cache_dir2, recursive = TRUE)

data_file <- file.path(tempdir(), "demo_data.tsv")
writeLines("gene\tcount\nTP53\t100\nBRCA1\t250", data_file)

read_data <- cacheFile(cache_dir2, backend = "rds") %@% function(path) {
  track_file(path)
  read.delim(path, sep = "\t")
}

summarize_data <- cacheFile(cache_dir2, backend = "rds") %@% function(path) {
  df <- read_data(path)
  setNames(df$count, df$gene)
}

print(summarize_data(data_file))

# Show tracked files
nodes <- cacheTree_nodes()
for (nid in names(nodes)) {
  nd <- nodes[[nid]]
  fh <- nd$file_hashes
  if (length(fh) > 0) {
    cat(sprintf("\n%s tracks %d file(s):\n", nd$fname, length(fh)))
    for (fp in names(fh)) {
      cat(sprintf("  %s  hash=%s...\n", fp, substr(fh[[fp]], 1, 16)))
    }
  }
}
```

## 4. Staleness Detection

`cache_tree_changed_files()` compares stored hashes against current disk state.

```{r}
# Before modification
stale <- cache_tree_changed_files()
cat(sprintf("Stale nodes before edit: %d\n", length(stale)))

# Modify the file
writeLines("gene\tcount\nTP53\t100\nBRCA1\t250\nEGFR\t500", data_file)

stale <- cache_tree_changed_files()
cat(sprintf("Stale nodes after edit:  %d\n", length(stale)))
for (nid in names(stale)) {
  cat(sprintf("  %s: %s\n", stale[[nid]]$node$fname,
      paste(stale[[nid]]$changed_files, collapse = ", ")))
}
```

## 5. Visualisation

`plot_cache_graph()` renders the DAG with igraph.

- **Navy** = cached and up-to-date
- **Amber** = stale (tracked file changed)
- **Blue** = tracked file node
- **Gray** = cache file missing

```{r}
if (requireNamespace("igraph", quietly = TRUE)) {
  plot_cache_graph()
  mtext("Cache Graph — after file modification", side = 3, line = -1.5, cex = 1.2, font = 2)
}
```

Save to file:

```{r, eval=FALSE}
plot_cache_graph(output = "my_graph.png")
```

## 6. Graph Persistence

The graph can be saved and loaded independently of the cache files.

```{r}
save_path <- file.path(tempdir(), "my_graph.rds")

# Save
cacheTree_save(save_path)
cat(sprintf("Saved %d nodes to %s\n", length(cacheTree_nodes()), save_path))

# Reset and verify
cacheTree_reset()
cat(sprintf("After reset: %d nodes\n", length(cacheTree_nodes())))

# Load back
cacheTree_load(save_path)
cat(sprintf("After load:  %d nodes\n", length(cacheTree_nodes())))

# Sync merges disk graph into memory (non-destructive)
cacheTree_reset()
cacheTree_sync(cache_dir2)
cat(sprintf("After sync:  %d nodes\n", length(cacheTree_nodes())))
```

## 7. Querying by File

`cacheTree_for_file(path)` finds all nodes that depend on a specific file.

```{r}
dependents <- cacheTree_for_file(normalizePath(data_file))
cat(sprintf("Nodes depending on %s:\n", basename(data_file)))
for (nid in names(dependents)) {
  cat(sprintf("  %s\n", dependents[[nid]]$fname))
}
```

## 8. Complex DAG

Build a diamond dependency: two branches merge into a final step.

```{r}
cacheTree_reset()
cache_dir3 <- file.path(tempdir(), "graph_demo_diamond")
if (dir.exists(cache_dir3)) unlink(cache_dir3, recursive = TRUE)

fetch_expression <- cacheFile(cache_dir3, backend = "rds") %@% function(sample) {
  c(TP53 = 10, BRCA1 = 25, EGFR = 8)
}

fetch_mutations <- cacheFile(cache_dir3, backend = "rds") %@% function(sample) {
  c(TP53 = TRUE, BRCA1 = FALSE, EGFR = TRUE)
}

branch_expression <- cacheFile(cache_dir3, backend = "rds") %@% function(sample) {
  expr <- fetch_expression(sample)
  expr[expr > 9]
}

branch_mutations <- cacheFile(cache_dir3, backend = "rds") %@% function(sample) {
  muts <- fetch_mutations(sample)
  names(muts[muts == TRUE])
}

integrate <- cacheFile(cache_dir3, backend = "rds") %@% function(sample) {
  high_expr <- branch_expression(sample)
  mutated <- branch_mutations(sample)
  high_expr[names(high_expr) %in% mutated]
}

result <- integrate("patient_001")
cat("Genes with high expression AND mutations:\n")
print(result)
```

```{r}
cat(sprintf("\nNodes: %d\n", length(cacheTree_nodes())))
for (nid in names(cacheTree_nodes())) {
  nd <- cacheTree_nodes()[[nid]]
  cat(sprintf("  %-25s  children=%d  parents=%d\n",
      nd$fname, length(nd$children), length(nd$parents)))
}
```

```{r}
if (requireNamespace("igraph", quietly = TRUE)) {
  plot_cache_graph()
  mtext("Diamond DAG — expression + mutation integration", side = 3, line = -1.5, cex = 1.2, font = 2)
}
```

## Cleanup

```{r}
unlink(c(cache_dir, cache_dir2, cache_dir3, data_file, save_path), recursive = TRUE)
```
