---
title: "Introduction to cacheR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to cacheR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

# Overview

`cacheR` provides **function-level caching** for R workflows.

Typical use cases in basic statistics:

- Expensive **summary statistics** on large datasets (mean, variance, quantiles).
- Repeated computation of **correlation matrices**.
- Multi-step pipelines, e.g. *preprocess → summarize → correlate*.

The main idea:

1. Wrap an existing function with `cacheFile()`.
2. Call the new function as usual.
3. `cacheR` saves results to disk and reuses them when inputs and function body have not changed.

---

# 1. Setup

```{r}
library(cacheR)
```

# If needed:
# library(decor)  # for the %@% operator

For this vignette we’ll use a **temporary cache directory**:

```{r}
cache_dir <- file.path(tempdir(), "cacheR-demo")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
cache_dir
```

---

# 2. Caching basic summary statistics

Assume we have a function that computes several statistics at once:

```{r}

stat_summary <- function(x) {
  message("Computing stats ...")
  Sys.sleep(1)  # pretend it's expensive
  x <- na.omit(x)
  list(
    mean     = mean(x),
    variance = var(x),
    sd       = sd(x),
    median   = median(x)
  )
}
```

We create a **cached** version:

```{r}

cached_stat_summary <- cacheFile(
  cache_dir = cache_dir,
  backend   = "rds"
) %@% function(x) {
  stat_summary(x)
}
```

Now, compare first vs second call:

```{r}

set.seed(1)
x <- rnorm(1e5)

system.time(res1 <- cached_stat_summary(x))
system.time(res2 <- cached_stat_summary(x))

res1
identical(res1, res2)
```

- First call: prints `"Computing stats ..."` and takes ~1 second.
- Second call with the *same* input: should be almost instant, reusing the cached result.

---

# 3. Caching a correlation matrix

Computing a full correlation matrix can be expensive for many variables.  
Let’s define a function that:

- standardizes the columns,
- computes the correlation matrix,
- optionally returns means and variances as well.

```{r}

cor_with_moments <- function(mat) {
  message("Computing correlations and moments ...")
  Sys.sleep(1)  # simulate expensive work

  # Basic checks
  stopifnot(is.matrix(mat) || is.data.frame(mat))
  mat <- as.matrix(mat)

  means     <- colMeans(mat, na.rm = TRUE)
  variances <- apply(mat, 2, var, na.rm = TRUE)

  corr_mat  <- stats::cor(mat, use = "pairwise.complete.obs")

  list(
    mean      = means,
    variance  = variances,
    cor       = corr_mat
  )
}
```

Wrap it with `cacheFile()`:

```{r}

cached_cor_with_moments <- cacheFile(
  cache_dir = cache_dir,
  backend   = "rds"
) %@% function(mat) {
  cor_with_moments(mat)
}
```

Test it:

```{r}

set.seed(123)
mat <- matrix(rnorm(200 * 10), ncol = 10)

system.time(res1 <- cached_cor_with_moments(mat))
system.time(res2 <- cached_cor_with_moments(mat))

str(res1$cor)
identical(res1$cor, res2$cor)
```

Again:

- First call: actually computes mean, variance and correlation.
- Second call: returns the result from disk, skipping the expensive computation.

---

# 4. Nested caching: reuse pre-processing

Suppose you often apply the same pre-processing (centering & scaling) before computing summary statistics or correlations.

Define a **preprocessing** function and a **downstream** function that uses it:

```{r}
preprocess_matrix <- function(mat) {
  message("Preprocessing matrix (center & scale) ...")
  Sys.sleep(1)  # simulate expensive step
  scale(mat)
}

# Cached preprocessing
cached_preprocess_matrix <- cacheFile(
  cache_dir = cache_dir
) %@% function(mat) {
  preprocess_matrix(mat)
}
```

Now a downstream function that uses the preprocessed data to compute means, variances and correlations:

```{r}
analyze_scaled <- cacheFile(
  cache_dir = cache_dir
) %@% function(mat) {
  message("Analyzing scaled data ...")
  mat_scaled <- cached_preprocess_matrix(mat)

  means     <- colMeans(mat_scaled)
  variances <- apply(mat_scaled, 2, var)
  corr_mat  <- stats::cor(mat_scaled)

  list(
    mean     = means,
    variance = variances,
    cor      = corr_mat
  )
}
```

Run it twice:

```{r}
set.seed(321)
mat2 <- matrix(rnorm(300 * 8), ncol = 8)

system.time(a1 <- analyze_scaled(mat2))
system.time(a2 <- analyze_scaled(mat2))

identical(a1$cor, a2$cor)
```

What happens:

1. **First run**
   - `cached_preprocess_matrix()` is executed and cached.
   - `analyze_scaled()` is executed and cached.
2. **Second run**
   - Both the preprocessing step and the full analysis should be loaded from cache
     (assuming the function bodies and input matrix are unchanged).

If your version of `cacheR` supports a **cache tree**, you can inspect the relationships between cached calls:

```{r}
cacheTree_reset()
a <- analyze_scaled(mat2)
nodes <- cacheTree_nodes()
nodes
```

You should see at least two nodes:

- one for `analyze_scaled(mat2)`
- one for `cached_preprocess_matrix(mat2)`

---

# 5. Using the `qs` backend for faster serialization

For large matrices, the [`qs`](https://github.com/qsbase/qs) backend can be faster than base RDS.

Install it once:

```{r, eval=FALSE}
install.packages("qs")
```

Then create a cached function with `backend = "qs"`:

```{r}
cached_cor_qs <- cacheFile(
  cache_dir = cache_dir,
  backend   = "qs"
) %@% function(mat) {
  cor_with_moments(mat)
}
```

Compare timings:

```{r}
set.seed(999)
mat_big <- matrix(rnorm(500 * 50), ncol = 50)

system.time(r1 <- cached_cor_qs(mat_big))
system.time(r2 <- cached_cor_qs(mat_big))

identical(r1$cor, r2$cor)
list.files(cache_dir, recursive = TRUE)
```

You should see some `.qs` files in the cache directory.

---

# 6. Combining file I/O and statistics

Create an example CSV file 

```{r make-example-csv, message=FALSE}
# We use tempdir() so that the vignette can be built anywhere
# without needing any external data files.

set.seed(1)

# Small numeric toy dataset
my_data <- data.frame(
  x = rnorm(100, mean = 0, sd = 1),
  y = rnorm(100, mean = 5, sd = 2),
  z = rnorm(100, mean = -2, sd = 0.5)
)

# Path where we'll save the CSV
example_csv <- file.path(tempdir(), "my_data.csv")

# Write the CSV (no row names to keep it clean)
write.csv(my_data, example_csv, row.names = FALSE)
```

In practice, you often:

1. Read data from a file (CSV, RDS, etc.),
2. Compute summary statistics or correlations.


You can cache both steps.

```{r}
read_numeric_csv <- function(path) {
  message("Reading CSV: ", path)
  Sys.sleep(1)  # simulate slow disk / network
  read.csv(path)
}

cached_read_numeric_csv <- cacheFile(
  cache_dir = cache_dir
) %@% function(path) {
  read_numeric_csv(path)
}

cached_file_stats <- cacheFile(
  cache_dir = cache_dir
) %@% function(path) {
  df <- cached_read_numeric_csv(path)
  cor_with_moments(df)
}
```

Usage:

```{r}
res <- cached_file_stats(example_csv)
res$mean
res$cor
```

- First run: reads the CSV and computes stats.
- Subsequent runs with the same file and same code: reuse cached results.

(Depending on your `cacheR` version, there may be options such as `file_args` to incorporate directory or file contents into the cache key more explicitly.)

---

# 7. Practical tips

- Use a project-local cache directory, e.g.:

  ```r
  cache_dir <- ".cacheR"
  dir.create(cache_dir, showWarnings = FALSE)
  ```

- Add it to `.gitignore`:

  ```text
  .cacheR/
  ```

- Wrap **expensive** operations:
  - reading large files
  - computing big correlation matrices
  - heavy preprocessing

- For pure statistical pipelines, a common pattern is:

  1. `cached_read()`         – load data
  2. `cached_preprocess()`   – clean / transform
  3. `cached_stats()`        – compute means, variances, correlations

  Each step can be shared across many analyses.

---

