---
title: "Accelerating ML Workflows with cacheR"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accelerating ML Workflows with cacheR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("randomForest", quietly = TRUE)
)
```

In Machine Learning pipelines, re-running expensive training steps just to tweak a plot or adjust a downstream report is a major productivity killer. cacheR solves this by automatically persisting your models based on their code, data, and hyperparameters.

In this example, we will simulate a training pipeline where we train multiple Random Forest models with different configurations and random seeds.

## 1. Setup

We will use the randomForest package. We also define a temporary directory for our cache.


```{r}
library(cacheR)
library(randomForest)
```

Setup a temporary cache location (in real projects, use a persistent path)

```{r}
cache_dir <- file.path(tempdir(), "ml_cache_example")
if(dir.exists(cache_dir)) unlink(cache_dir, recursive = TRUE)
```


## 2. Define the Cached Trainer

We wrap our training logic with cacheFile.

Key Pattern: Notice that seed is passed as an argument to the function. This ensures that:

The seed is part of the cache hash (changing the seed invalidates the cache).

The function body sets the seed, guaranteeing deterministic results for that specific entry.

We add a generic 1-second sleep to simulate a large dataset or complex training

```{r}
train_rf <- cacheFile(cache_dir) %@% function(data, ntree, mtry, seed) {
  message(sprintf("⚡ Training Model: ntree=%d, mtry=%d, seed=%d ...", ntree, mtry, seed))
  
  # Ensure reproducibility inside the worker
  set.seed(seed)
  
  # Simulate expensive computation
  Sys.sleep(1) 
  
  # Actual training
  model <- randomForest(Species ~ ., data = data, ntree = ntree, mtry = mtry)
  return(model)
}
```

## 3. Experiment: Cold Cache vs. Warm Cache
We will now train three distinct model variations.

Phase A: The "Cold" Run (Training)
These calls will take time because the cache is empty. We track the start time to measure performance.

```{r}
data(iris)

cat("--- PHASE A: Initial Training (Cold Cache) ---\n")
start_time <- Sys.time()

# Model 1: Base configuration
model_1 <- train_rf(data = iris, ntree = 500, mtry = 2, seed = 42)

# Model 2: Different Hyperparameter (ntree) -> New Cache Entry
model_2 <- train_rf(data = iris, ntree = 1000, mtry = 2, seed = 42)

# Model 3: Same parameters, DIFFERENT seed -> New Cache Entry
model_3 <- train_rf(data = iris, ntree = 500, mtry = 2, seed = 99)

time_cold <- Sys.time() - start_time
cat(sprintf("\nTotal Training Time: %.2f seconds\n", as.numeric(time_cold, units="secs")))
```

Phase B: The "Warm" Run (Retrieval)
Now, imagine you restarted your R session or are simply knitting an RMarkdown report. You run the exact same code. cacheR detects that the inputs (and the function body) haven't changed.

```{r}

cat("\n--- PHASE B: Reloading Models (Warm Cache) ---\n")
start_time <- Sys.time()

# Re-requesting the exact same models
model_1_cached <- train_rf(data = iris, ntree = 500, mtry = 2, seed = 42)
model_2_cached <- train_rf(data = iris, ntree = 1000, mtry = 2, seed = 42)
model_3_cached <- train_rf(data = iris, ntree = 500, mtry = 2, seed = 99)

time_warm <- Sys.time() - start_time
cat(sprintf("\nTotal Retrieval Time: %.2f seconds\n", as.numeric(time_warm, units="secs")))
```


Phase C: Verification
We can verify that the models are indeed identical. Because we tracked the seed in the cache key, model_1 and model_3 (which differed only by seed) are correctly stored as separate files.

```{r}
# Verify Model 1 matches its cached version
identical(model_1$importance, model_1_cached$importance) 
# [1] TRUE

# Verify Model 1 (Seed 42) is different from Model 3 (Seed 99)
identical(model_1$importance, model_3$importance) 
# [1] FALSE
```