---
title: "Reproducible and Fast RNA-seq Analysis with cacheR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Reproducible and Fast RNA-seq Analysis with cacheR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# cacheR RNA-seq Caching Example 

This document describes how to use the `cacheR` package to cache expensive steps in an RNA-seq data analysis pipeline. It focuses on: 

- Importing RNA-seq count files from disk 

- Building a `DESeqDataSet`

- Fitting a DESeq model

- Automatically invalidating caches when:

- Function arguments change
- Function code changes 
- The number of files in key directories changes
- Hardcoded and global-variable paths change The    examples assume you are working inside an R package or project that includes the `cacheFile()` decorator from `cacheR`.

--- 

## 1. Setup 

### 1.1 Required packages You will need: 

- `cacheR`  for caching and dependency tracking 
- `DESeq2`  for differential expression analysis 
- `dplyr`, `readr`  for data handling 

```{r, echo = FALSE} 
suppressPackageStartupMessages({
        library(cacheR) 
        library(DESeq2) 
        library(dplyr) 
        library(readr) 
})
``` 

### 1.2 Working directories We use a temporary working directory for demonstration. In a real project, you would use a project-specific path.

```{r} 
work_dir <- file.path(tempdir(), "cacheR_rnaseq_vignette") 
dir.create(work_dir, recursive = TRUE, showWarnings = FALSE) 

counts_dir <- file.path(work_dir, "counts") 
dir.create(counts_dir, recursive = TRUE, showWarnings = FALSE) 

cache_dir <- file.path(work_dir, "cache") 
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
```

---

## 2. Simulated RNA-seq data To keep the example self-contained, we simulate small count tables as if they were generated by tools such as `featureCounts` or `htseq-count`.

```{r} 

set.seed(1) 

genes <- paste0("Gene", sprintf("%03d", 1:500)) 

simulate_sample <- function(sample_id, n_genes = length(genes)) { 
    data.frame( 
        gene_id = genes, 
        count = rpois(n_genes, lambda = sample(10:1000, 1)), stringsAsFactors = FALSE ) 
} 
        
sample_ids <- paste0("sample", 1:4) 

for (sid in sample_ids) { 
    tab <- simulate_sample(sid) 
    readr::write_tsv(tab, file.path(counts_dir, paste0(sid, "_counts.tsv")) ) 
} 
sample_table <- data.frame( sample_id = sample_ids, condition = rep(c("A", "B"), each = 2), stringsAsFactors = FALSE )

``` 

At this point: - `counts_dir` contains 4 tab-separated count files, one per sample. - `sample_table` describes the experimental design (two conditions: A and B).

--- 

## 3. Reading count files from a directory 

### 3.1 Uncached helper function 

The following function reads all `*_counts.tsv` files from a directory, merges them on `gene_id`, and orders columns to match the sample table.

```{r} 

read_counts_from_dir <- function(count_dir, sample_table) { 
files <- list.files(count_dir, pattern = "_counts\\.tsv$", full.names = TRUE) 
if (length(files) == 0L) { 
        stop("No count files found in: ", count_dir) 
} 
sample_ids <- gsub("_counts\\.tsv$", "", basename(files)) 
counts_list <- lapply(files, function(f) { 
        readr::read_tsv(f, show_col_types = FALSE) 
}) 
merged <- counts_list[[1]][, "gene_id", drop = FALSE] 
for (i in seq_along(counts_list)) { 
        merged[[sample_ids[i]]] <- counts_list[[i]]$count } 
        merged <- merged[, c("gene_id", sample_table$sample_id), drop = FALSE] 
        merged 
}

``` 

### 3.2 Caching the file import step 

We now wrap `read_counts_from_dir()` with `cacheFile()`. The key point is the argument `file_args = "count_dir"`: - `cacheR` will include the **number of files** in `count_dir` as part of the cache key. - If new files are added or removed, the hash changes and the function is re-run.

```{r} 
cached_read_counts <- cacheR:::cacheFile( cache_dir = cache_dir ) %@% function(count_dir, sample_table) { 
    read_counts_from_dir(count_dir, sample_table) 
}

``` 

### 3.3 First and second run

```{r}

# First run: reads from disk and caches 

```{r}
counts1 <- cached_read_counts(counts_dir, sample_table)
```

# Second run: same arguments and same files → uses cache 

```{r, eval=TRUE}
counts2 <- cached_read_counts(counts_dir, sample_table) 

identical(counts1, counts2)
```

should be **TRUE**

---

## 4. Automatic invalidation when new samples appear 

If you add a new sample and its counts file to `counts_dir`, the cache should be invalidated because the number of files in `count_dir` changed.

```{r} 
new_id <- "sample5" 
sample_table2 <- bind_rows( sample_table, data.frame(sample_id = new_id, condition = "A") ) 
tab_new <- simulate_sample(new_id) 
readr::write_tsv(tab_new, file.path(counts_dir, paste0(new_id, "_counts.tsv")) ) 
counts3 <- cached_read_counts(counts_dir, sample_table2) 

dim(counts1)
```

dimensions before adding sample5 ```dim(counts3)```

dimensions after adding sample5

Key behavior: - `cached_read_counts()` detects that `counts_dir` now has 5 files instead of 4. - The cache key changes, so the function is re-run. - `counts3` now contains the additional `sample5` column. 

--- 

## 5. Building a DESeqDataSet with caching 

Constructing a `DESeqDataSet` can be computationally expensive for large matrices. We can cache this step as well. 

### 5.1 Cached DESeqDataSet construction

```{r} 
dds_from_counts <- cacheR:::cacheFile( cache_dir = cache_dir ) %@% function(counts_tbl, sample_table) { 
       
        mat <- as.matrix(counts_tbl[, sample_table$sample_id]) 
        
        rownames(mat) <- counts_tbl$gene_id 
        
        coldata <- as.data.frame(sample_table) 
        
        rownames(coldata) <- coldata$sample_id

        DESeqDataSetFromMatrix( 
                countData = mat, 
                colData   = coldata, 
                design    = ~condition 
        ) 
        
}


``` 

### 5.2 Usage

```{r} 

dds <- dds_from_counts(counts3, sample_table2)
dds

# standard DESeq2 object

``` 

Re-running `dds_from_counts(counts3, sample_table2)` with unchanged inputs will load from the cache instead of recomputing. 

--- 

## 6. Caching the DESeq model fit 

Fitting the DESeq model is often the slowest step in the pipeline. 

### 6.1 Cached DESeq fit

```{r} 

run_deseq <- cacheR:::cacheFile( cache_dir = cache_dir ) %@% function(dds) { 
        DESeq(dds) 
        } 
dds_fit <- run_deseq(dds)

``` 

If you run `run_deseq(dds)` again with the same `dds`, it will use the cached result rather than re-fitting the model.

---

### 6.2 Extracting results Extracting results from the fitted model is typically fast, so you may or may not want to cache it:

```{r} 

res <- results(dds_fit, contrast = c("condition", "B", "A")) 

head(res[order(res$padj), ])

``` 

---

## 7. Tracking hardcoded directories and global variables 

`cacheR` can also track directories that are not passed as function arguments, but appear in: 

- Hardcoded calls like `list.files("/path/to/data")` 

- Global variable calls like `dir(GLOBAL_DIR)`

### 7.1 Global directory example

```{r} 

GLOBAL_DIR <- counts_dir 

hardcoded_fun <- cacheR:::cacheFile( cache_dir = cache_dir ) %@% function() {

        # Using a global variable as directory path 
        n1 <- length(list.files(GLOBAL_DIR)) 
        n2 <- length(dir(GLOBAL_DIR)) 
        c(literal = n1, global = n2) 
}

# First call: computes and caches 

hardcoded_fun()

# Add another file to the directory 

extra <- simulate_sample("extra") 

readr::write_tsv(extra, file.path(GLOBAL_DIR, "extra_counts.tsv") )

# Second call: cache is invalidated and recomputed 

hardcoded_fun()

``` 

Here: - The function body uses `GLOBAL_DIR` inside `list.files()` and `dir()`. - `cacheR` inspects the function body, detects these directory usages, and includes the corresponding file counts in the cache key. - When a new file is added to `GLOBAL_DIR`, the cache entry becomes invalid, and `hardcoded_fun()` is re-run. 

---

## 8. Practical notes 

**Where to store cache files** 

Use a project-specific directory, for example `cache/` in your project root, or a configurable location via `options(cacheR.dir = ...)`. 

**Choosing what to cache** 

Good candidates are: 

- Disk I/O heavy operations (reading many files) 

- Large object construction (`DESeqDataSet`, `SummarizedExperiment`, etc.) 

- Expensive model fitting steps - 

**What not to cache** 

Very fast operations (simple filtering, small transforms) rarely benefit from caching and can add overhead. 

---


## Session Info

```{r}
sessionInfo()
```